$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-09-22 paper/paper.tex
a pass on technical parts

\begin{proof}[Proof (Sketch)]
  All the above results are proved by induction on \textsc{Process} and \textsc{Var} (\autoref{fig:syntax}).
  Complete details can be found in our mechanisation in Agda%
\begin{anonsuppress}
in \cite{Zalakain2020Agda}
\end{anonsuppress}
.
\end{proof}

@@

\begin{nidefinition}[Usage algebra]
  A \emph{usage algebra} is a ternary relation $\op{x}{y}{z}$ that is \emph{partial} (as not any two multiplicities can be combined), \emph{deterministic} and \emph{cancellative} (to aid equational reasoning), \emph{associative} and \emph{commutative} (following directly from subject congruence for parallel composition), and in which the leftovers can be \emph{computed} (as to automatically update the usage context every time input and output occurs).
  It has a \emph{neutral element} $\zero$ that is absorbed on either side, and that is also \emph{minimal} (so that new resources cannot arbitrarily spring into life).
  It has an element $\one$ that is used to count inputs and outputs.
  \autoref{fig:multiplicities} models such an algebra as a record $\Algebra_C$ on a carrier $C$.

@@

\begin{nidefinition}[\textsc{Type} and \textsc{PreCtx}: types and typing contexts]
  A \emph{type} is either a unit type ($\unit$), a base type ($\base{n}$) or a channel type ($\channel{t}{x}$) (\autoref{fig:prectx}).

@@

        \inferrule
            {\gamma : \PreCtx_n \\ t : \Type}
            {\gamma \comma t : \PreCtx_{\suc n}}
  \end{mathpar}
  \caption{Types and length-indexed typing contexts.}
  \label{fig:prectx}
  \end{figure}
\end{nidefinition}

The unit type $\unit$ serves as a proof of inhabitance for types.
The base type $\base{n}$ uses natural numbers as placeholders for types (the host language can then interpret the former into the latter --- e.g., $0$ as booleans, $1$ as natural numbers) so that we avoid having to deal with universe polymorphism.
The type $\channel{t}{x}$ of a channel determines what type ${t}$ of data and what usage annotations ${x}$ are sent over that channel.
This channel notation aligns with $[t]\ \mathbf{chan}_{(i^{y},o^{z})}$, where $y,z$ are the multiplicities of input and output, respectively \cite{K07}.

A \emph{typing context} (last row in \autoref{fig:prectx}) is a length-indexed list of types, and is either empty $[]$ or the result of appending a type $t$ to an existing context $\gamma$.

@@

  We use the notation $\type{C}^{\func{2}}$ to stand for a $\type{C} \constr{\times} \type{C}$ tuple of input and output multiplicities, respectively.
  Henceforth, we use $\lz$ to denote the pair $0 \comma 0$ for a channel which cannot be used any further, $\li$ for the pair $1 \comma 0$ for a channel to be used in input exactly once, $\lo$ for $0 \comma 1$ for a channel to be used in output exactly once, and $\lio$ for $1 \comma 1$, for a channel to be used exactly once in input and exactly once in output.
  This notation was originally used in the linear \picalc{} \cite{KPT96,Sangio01}.

@@

The \receiver{} defined by the $\func{recv}$ process receives data along the channel with index $0$, thus that variable needs to be of type $\channel{t}{u}$ for some $t$ and $u$.
After receiving twice, the process ends, and we should not be left with any unused multiplicities: $u$ needs to be $\lz$.
We will use graded types to keep track of the exact number of times communication happens.
Whatever the input multiplicity of the channel, we will consume $2$ of it and leave the remaining as leftovers.
Once the types are given, the typing derivation is completely syntax directed.

The \sender{} defined by the $\func{send}$ process sends data along the channel with index $0$, thus that variable needs to be of type $\channel{t}{u}$ for some $t$ and $u$.
We instantiate $t$ (the type of data that we transmit) to $\channel{\unit}{\func{\omega}}$, as that is a value we can easily construct.
As per the type of the process $\func{recv}$, the transmitted multiplicities $u$ are $\lz$.
We will transmit once, thus using a single output multiplicity, and leave the rest as leftovers.

\begin{flalign*}
  & \func{\vdash recv} \; \types{\gamma \comma \channel{t}{\lz}}{\Gamma \comma (\suc \suc l \comma r)}{\func{recv} \; \constr{0}}{\Gamma \comma (l \comma r)} & \\
  & \func{\vdash recv} \; = \; \recv{\constr{0}}{} \recv{(\suc \constr{0})}{} \PO \\
  & \\
  & \func{\vdash send} \; \types{\gamma \comma \channel{\channel{\unit}{\func{\omega}}}{\lz}}{\Gamma \comma (l \comma \suc r)}{\func{send} \; \constr{0}}{\Gamma \comma (l \comma r)} & \\
  & \func{\vdash send} \; = \; \new{} \; \_ \; \_ \; \zero \; \send{(\suc \constr{0})}{\constr{0}} \PO \\
\end{flalign*}

Dually, the \courier{} defined by the $\func{carry}$ process expects input multiplicities for the channels shared with $\func{send}$ and output multiplicities for the channel shared with $\func{recv}$.
The typing derivation here is uniquely determined by the type and syntax of the process.

\begin{flalign*}
& \func{\vdash carry} && : \types{\gamma \comma \channel{t}{\lz} \comma \channel{t}{\lz} \comma \channel{t}{\lz}\\ & &&}{\Gamma \comma (\suc lx \comma rx) \comma (\suc ly \comma ry) \comma (lz \comma \suc \suc rz)\\ & &&}{\func{carry} \; (\suc \suc \constr{0}) \; (\suc \constr{0}) \; \constr{0}\\ & &&}{\Gamma \comma (lx \comma rx) \comma (ly \comma ry) \comma (lz \comma rz)} & \\
& \func{\vdash carry} && = \\
& && \recv{(\suc \suc \constr{0})}{} \\
& && \recv{(\suc \suc \constr{0})}{} \\
& && \send{(\suc \suc \constr{0})}{\suc \constr{0}} \\
& && \send{(\suc \suc \constr{0})}{\constr{0}} \PO \\
\end{flalign*}

We can now put these processes together to type the courier system.
Specifying the types of the channels is not required: Agda can infer them from the types of the processes.

\begin{flalign*}
& \func{\vdash system} && : \; \types{\constr{[]}}{\constr{[]}}{\func{system}}{\constr{[]}} & \\
& \func{\vdash system} && = \new{} \; \_ \; \_ \; \_ \; ( \func{\vdash send} \\
& && \comp{}{ \new{} \; \_ \; \_ \; \_ } \; ( \func{\vdash send} \\
& && \comp{}{ \new{} \; \_ \; \_ \; \_ } \; ( \func{\vdash recv} \\
& && \comp{}{ \func{\vdash carry} }))) \\
\end{flalign*}

\end{example}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Type Safety}
\label{type-safety}
\subsection{Auxiliary Results}

\paragraph*{Framing}
asserts that the well-typedness of a process is independent of its leftover resources.

\begin{nitheorem}[Framing]
  \label{thm:framing}
  Let $P$ be well typed in $\types{\gamma}{\Gamma_l}{P}{\Xi_l}$.
  Let $\Delta$ be such that $\opctx{\Gamma_l}{\Delta}{\Xi_l}$.
  Let $\Gamma_r$ and $\Xi_r$ be arbitrary contexts such that $\opctx{\Gamma_r}{\Delta}{\Xi_r}$.

  Then $\types{\gamma}{\Gamma_r}{P}{\Xi_r}$.
\end{nitheorem}

\paragraph*{Weakening}
states that inserting a new variable into the context preserves the well-typedness of a process as long as the usage annotation of the inserted variable is preserved as a leftover.
Let $\func{ins}_i$ insert an element into a context at position $i$ --- for simplicity, we use it both to insert types into typing contexts and usage annotations into usage contexts.
\begin{nitheorem}[Weakening]
  \label{thm:weakening}
  Let $P$ be well typed in $\types{\gamma}{\Gamma}{P}{\Xi}$.
  Then, lifting every variable greater than or equal to $i$ in $P$ is well typed in
  $\types{\func{ins}_i \; t \; \gamma}{\func{ins}_i \; x \; \Gamma}{\func{lift}_i \; P}{\func{ins}_i \; x \; \Xi}$.
\end{nitheorem}

\paragraph*{Strengthening}
states that removing an unused variable preserves the well-typedness of a process.
Let $\func{del}_i$ delete the element at position $i$ from a context --- for simplicity, we use it both to delete types from typing contexts and usage annotations from usage contexts.
\begin{nitheorem}[Strengthening]
  \label{thm:strengthening} 
  Let $P$ be well typed in $\types{\gamma}{\Gamma}{P}{\Xi}$.
  Let $i$ be a variable not in $P$, such that $\Unused_i \; P$.
  Then lowering every variable greater than $i$ in $P$ is well typed in $\types{\func{del}_i \; \gamma}{\func{del}_i \; \Gamma}{\func{lower}_i \; P}{\func{del}_i \; \Xi}$.
\end{nitheorem}

% \begin{remark}
  The \emph{unusedness} predicate cannot depend on the preservation of usage annotations by a process: the process could preserve such annotations, yet sent a $\lz$ usage of those variables over a channel.
% \end{remark}

\paragraph*{Exchange}
of two variables preserves the well-typedness of a process.
We extend $\func{exchange}_i$  introduced in \autoref{structural-congruence} to exchange types in typing contexts and usage annotations in usage contexts.
\begin{nitheorem}[Exchange]
  \label{thm:exchange}
  Let $P$ be well typed in $\types{\gamma}{\Gamma}{P}{\Xi}$.
  Then, $\types{\func{exchange}_i \; \gamma}{\func{exchange}_i \; \Gamma}{\allowbreak \func{exchange}_i \; P}{\func{exchange}_i \; \Xi}$.
\end{nitheorem}

\begin{proof}[Proof (Sketch)]
  All the above theorems are proved by induction on \textsc{Types} and \textsc{VarRef}.
  For details, refer to our mechanisation in Agda%
\begin{anonsuppress}
\cite{Zalakain2020Agda}
\end{anonsuppress}
.
\end{proof}

\subsection{Towards Subject Reduction}
\paragraph*{Subject Congruence}
states that applying structural congruence rules to a well typed process preserves its well-typedness.
We must first introduce lemmas that establish that some syntactic manipulations can be inverted and how unused variables relate to the preservation of leftovers.

@@

\begin{proof}[Proof (Sketch)]
  The proof is by induction on \textsc{Equals} $\eq{}$.
  Here we only consider those cases that are not purely inductive: the base cases for $\constr{struct}$ and their symmetric variants.
\begin{anonsuppress}
Full proof in \cite{Zalakain2020Agda}.
\end{anonsuppress}
  We proceed by induction on \textsc{StructCong} $\eqeq$:
  \begin{itemize}
    \item
      Case $\constr{comp-assoc}$: trivial, as leftover typing is naturally associative.
    \item
      Case $\constr{comp-sym}$ for $\comp{P}{Q}$: we use framing (\autoref{thm:framing}) to shift the output context of $P$ to the one of $Q$; and the input context of $Q$ to the one of $P$.
    \item
      Case $\constr{comp-end}$: trivial, as the typing rule for $\PO$ has the same input and output contexts.
    \item
      Case $\constr{scope-end}$: we show that the usage annotation of the newly created channel must be $\lz$, making the proof trivial.
      In the opposite direction, we instantiate the newly created channel to a type $\unit$ and a usage annotation $\lz$.
    \item
      Case $\constr{scope-ext}$ for $\new \comp{P}{Q}$: we to show that $P$ preserves the usage annotations of the unused variable (\autoref{lm:types-unused}) and then use strengthening (\autoref{thm:strengthening}).
      In the reverse direction, we use weakening (\autoref{thm:weakening}) on $P$ and show that lowering and then lifting $P$ results in $P$ (\autoref{lm:lower-lift}).
    \item
      Case $\constr{scope-comm}$:
      we use exchange (\autoref{thm:exchange}), and for the reverse direction exchange and \autoref{lm:exchange-exchange} to show that exchanging two elements in $P$ twice leaves $P$ unchanged. \qedhere
  \end{itemize}
\end{proof}

\paragraph*{Substitution}
\autoref{thm:substitution-generalization} proves a generalised version of substitution, where the substitition $\subst{P}{j}{i}$ is on any variable $i$.
\autoref{thm:substitution} instantiates the generalised version to the concrete case where $i$ is the most recently introduced variable $\constr{0}$, as required by subject reduction.

@@

\begin{proof}[Proof (Sketch)]
  By induction on $\types{\gamma}{\Gamma_i}{P}{\Psi_i}$.
  Full details are found in \autoref{app:substitution-generalization}.
  \begin{itemize}
   \item
     Constructor $\PO$: observe that $\Gamma_i \equiv \Psi_i$ and thus $\Gamma_j \equiv \Psi_j$.
   \item
     Constructor $\new$: we proceed inductively.
   \item
     Constructors $\recv{}{}$, $\send{}{}$ and $\comp{}{}$: we must find $\Theta$ in \autoref{fig:subst} and split the diagram along its vertical axis to proceed by induction. \qedhere
 \end{itemize}
\end{proof}

@@

\begin{nitheorem}[Substitution]
  \label{thm:substitution}
  Let $P$ be well typed in $\types{\gamma \comma t}{\Gamma \comma m}{P}{\Psi \comma \lz}$.
  Let $\contains{\gamma}{\Psi}{j}{t}{m}{\Xi}$.
  Then, we can substitute the variable references to $\constr{0}$ in $P$ with $\suc j$ so that the result is well typed in $\types{\gamma \comma t}{\Gamma \comma m}{\subst{P}{\suc j}{\constr{0}}}{\Xi \comma m}$.
\end{nitheorem}
\begin{proof}[Proof (Sketch)]
  We use framing to derive $\contains{\gamma}{\Gamma}{j}{t}{m}{\Theta}$ and $\types{\gamma \comma t}{\Theta \comma m}{P}{\Xi \comma \lz}$ for some $\Theta$.
  Then, we use these to apply \autoref{thm:substitution-generalization}.
\end{proof}

@@

\paragraph*{Subject Reduction}
states that if $P$ is well typed and it reduces to $Q$, then $Q$ is well typed.
In the \picalc{} we distinguish between a reduction $P \reduce{\constr{internal}} Q$ on a channel internal to $P$, and a reduction $P \reduce{\constr{external} \; i} Q$ on a channel $i$ external to $P$ (refer to \autoref{operational-semantics}).
We first introduce a helper lemma:

@@

\begin{proof}[Proof (Sketch)]
  By induction on $P \reduce{c} Q$.
  \hfill{}\\
  \begin{itemize}
    \item
    Case $\constr{comm}$: we apply framing (\autoref{thm:framing}) (to rearrange the assumptions), substitution (\autoref{thm:substitution}) and strengthening (\autoref{thm:strengthening}).

>>>>

\begin{proof}[Proof]
  All the above results are proved by induction on \textsc{Process} and \textsc{Var} (\autoref{fig:syntax}).
  Complete details can be found in our mechanisation in Agda%
\begin{anonsuppress}
in \cite{Zalakain2020Agda}
\end{anonsuppress}
.
\end{proof}

@@

\begin{nidefinition}[Usage algebra]
  A \emph{usage algebra} is a ternary relation $\op{x}{y}{z}$ that is \emph{partial} (as not any two multiplicities can be combined), \emph{deterministic} and \emph{cancellative} (to aid equational reasoning), \emph{associative} and \emph{commutative} (following directly from subject congruence for parallel composition), and in which the leftovers can be \emph{computed} (as to automatically update the usage context every time input and output occurs).
  It has a \emph{neutral element} $\zero$ that is absorbed on either side, and that is also \emph{minimal} (so that new resources cannot arbitrarily spring into life).
  It has an element $\one$ that is used to count inputs and outputs.
  \autoref{fig:multiplicities} defines such an algebra as a record $\Algebra_C$ on a carrier $C$.

@@

\begin{nidefinition}[\textsc{Type} and \textsc{PreCtx}: types and typing contexts]
  A \emph{type} is either a unit type ($\unit$), a base type ($\base{n}$) or a channel type ($\channel{t}{x}$) (first two rows of \autoref{fig:prectx}).

@@

        \inferrule
            {\gamma : \PreCtx_n \\ t : \Type}
            {\gamma \comma t : \PreCtx_{\suc n}}
  \end{mathpar}
  \caption{Types and length-indexed typing contexts.}
  \label{fig:prectx}
  \end{figure}
The unit type $\unit$ serves as a proof of inhabitance for types.
The base type $\base{n}$ uses natural numbers as placeholders for types (the host language can then interpret the former into the latter --- e.g., $0$ as booleans, $1$ as natural numbers) so that we avoid having to deal with universe polymorphism.
The type $\channel{t}{x}$ of a channel determines what type ${t}$ of data and what usage annotations ${x}$ are sent over that channel.
This channel notation aligns with $[t]\ \mathbf{chan}_{(i^{y},o^{z})}$, where $y,z$ are the multiplicities of input $i$ and output $o$ capabilities, respectively \cite{K07}.

A \emph{typing context} (last two rows of \autoref{fig:prectx}) is a length-indexed list of types, and is either empty $[]$ or the result of appending a type $t$ to an existing context $\gamma$.
\end{nidefinition}

@@

  We use the notation $\type{C}^{\func{2}}$ to stand for a $\type{C} \constr{\times} \type{C}$ pair of input and output multiplicities, respectively.
  Henceforth, we use $\lz$ to denote the pair $0 \comma 0$ for a channel which cannot be used any further, $\li$ for the pair $1 \comma 0$ for a channel to be used in input exactly once, $\lo$ for $0 \comma 1$ for a channel to be used in output exactly once, and $\lio$ for $1 \comma 1$, for a channel to be used exactly once in input and exactly once in output.
  This notation was originally used in the linear \picalc{} \cite{KPT96,Sangio01}.

@@

The \receiver{} defined by the $\func{recv}$ process receives data along the channel with index $0$, thus that variable needs to be of channel type $\channel{t}{u}$ for some $t$ and $u$.
After receiving twice, the process ends, and we should not be left with any unused multiplicities: $u$ needs to be $\lz$.
We will use graded types to keep track of the exact number of times communication happens.
Whatever the input multiplicity of the channel, we will consume $2$ of it and leave the remaining as leftovers.
Once the types are given, the typing derivation is completely syntax directed.
\begin{flalign*}
  & \func{\vdash recv} \; \types{\gamma \comma \channel{t}{\lz}}{\Gamma \comma (\suc \suc l \comma r)}{\func{recv} \; \constr{0}}{\Gamma \comma (l \comma r)} & \\
  & \func{\vdash recv} \; = \; \recv{\constr{0}}{} \recv{(\suc \constr{0})}{} \PO
\end{flalign*}

The \sender{} defined by the $\func{send}$ process sends data along the channel with index $0$, thus that variable needs to be of channel type $\channel{t}{u}$ for some $t$ and $u$.
We instantiate $t$ (the type of data that the \sender{} sends) to $\channel{\unit}{\func{\omega}}$, as that is a value we can easily construct.
As per the type of the process $\func{recv}$, the transmitted multiplicities $u$ are $\lz$.
We will transmit once, thus using a single output multiplicity, and leave the rest as leftovers.
\begin{flalign*}
  & \func{\vdash send} \; \types{\gamma \comma \channel{\channel{\unit}{\func{\omega}}}{\lz}}{\Gamma \comma (l \comma \suc r)}{\func{send} \; \constr{0}}{\Gamma \comma (l \comma r)} & \\
  & \func{\vdash send} \; = \; \new{} \; \_ \; \_ \; \zero \; \send{(\suc \constr{0})}{\constr{0}} \PO
\end{flalign*}

Dually, the \courier{} defined by the $\func{carry}$ process expects input multiplicities for the channels shared with $\func{send}$ and output multiplicities for the channel shared with $\func{recv}$.
The typing derivation here is uniquely determined by the type and syntax of the process.
\begin{flalign*}
& \func{\vdash carry} && : \types{\gamma \comma \channel{t}{\lz} \comma \channel{t}{\lz} \comma \channel{t}{\lz}\\ & &&}{\Gamma \comma (\suc lx \comma rx) \comma (\suc ly \comma ry) \comma (lz \comma \suc \suc rz)\\ & &&}{\func{carry} \; (\suc \suc \constr{0}) \; (\suc \constr{0}) \; \constr{0}\\ & &&}{\Gamma \comma (lx \comma rx) \comma (ly \comma ry) \comma (lz \comma rz)} & \\
& \func{\vdash carry} && = \\
& && \recv{(\suc \suc \constr{0})}{} \\
& && \recv{(\suc \suc \constr{0})}{} \\
& && \send{(\suc \suc \constr{0})}{\suc \constr{0}} \\
& && \send{(\suc \suc \constr{0})}{\constr{0}} \PO
\end{flalign*}

We can now compose these processes in parallel and type the courier system.
Specifying the types of the channels is not required: Agda can infer them from the types of the processes.
\begin{flalign*}
& \func{\vdash system} && : \; \types{\constr{[]}}{\constr{[]}}{\func{system}}{\constr{[]}} & \\
& \func{\vdash system} && = \new{} \; \_ \; \_ \; \_ \; ( \func{\vdash send} \\
& && \comp{}{ \new{} \; \_ \; \_ \; \_ } \; ( \func{\vdash send} \\
& && \comp{}{ \new{} \; \_ \; \_ \; \_ } \; ( \func{\vdash recv} \\
& && \comp{}{ \func{\vdash carry} })))
\end{flalign*}
\end{example}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Type Safety}
\label{type-safety}
\subsection{Auxiliary Results}
We will start this section with auxiliary results which are needed to prove two main theorems: subject congruence (\autoref{thm:subject-congruence}) and subject reduction (\autoref{thm:subject-reduction}) for our \picalc{} with leftovers.

\paragraph*{Framing}
This property asserts that the well-typedness of a process is independent of its leftover resources.

\begin{nitheorem}[Framing]
  \label{thm:framing}
  Let $P$ be well typed in $\types{\gamma}{\Gamma_l}{P}{\Xi_l}$.
  Let $\Delta$ be such that $\opctx{\Gamma_l}{\Delta}{\Xi_l}$.
  Let $\Gamma_r$ and $\Xi_r$ be arbitrary contexts such that $\opctx{\Gamma_r}{\Delta}{\Xi_r}$.
  Then $\types{\gamma}{\Gamma_r}{P}{\Xi_r}$.
\end{nitheorem}

\paragraph*{Weakening}
This property states that inserting a new variable into the context preserves the well-typedness of a process as long as the usage annotation of the inserted variable is preserved as a leftover.
Let $\func{ins}_i$ insert an element into a context at position $i$ --- for simplicity, we use it both to insert types into typing contexts and usage annotations into usage contexts.
\begin{nitheorem}[Weakening]
  \label{thm:weakening}
  Let $P$ be well typed in $\types{\gamma}{\Gamma}{P}{\Xi}$.
  Then, lifting every variable greater than or equal to $i$ in $P$ is well typed in
  $\types{\func{ins}_i \; t \; \gamma}{\func{ins}_i \; x \; \Gamma}{\func{lift}_i \; P}{\func{ins}_i \; x \; \Xi}$.
\end{nitheorem}

\paragraph*{Strengthening}
This property states that removing an unused variable preserves the well-typedness of a process.
Let $\func{del}_i$ delete the element at position $i$ from a context --- for simplicity, we use it both to delete types from typing contexts and usage annotations from usage contexts.
\begin{nitheorem}[Strengthening]
  \label{thm:strengthening} 
  Let $P$ be well typed in $\types{\gamma}{\Gamma}{P}{\Xi}$.
  Let $i$ be a variable not in $P$, such that $\Unused_i \; P$.
  Then lowering every variable greater than $i$ in $P$ is well typed in $\types{\func{del}_i \; \gamma}{\func{del}_i \; \Gamma}{\func{lower}_i \; P}{\func{del}_i \; \Xi}$.
\end{nitheorem}

% \begin{remark}
  Notice that the \emph{unused} predicate cannot depend on the preservation of usage annotations by a process: the process could preserve such annotations, yet sent a $\lz$ usage of those variables over a channel.
% \end{remark}

\paragraph*{Exchange}
This property states that the exchange of two variables preserves the well-typedness of a process.
We extend $\func{exchange}_i$  introduced in \autoref{structural-congruence} to exchange types in typing contexts and usage annotations in usage contexts.
\begin{nitheorem}[Exchange]
  \label{thm:exchange}
  Let $P$ be well typed in $\types{\gamma}{\Gamma}{P}{\Xi}$.
  Then, $\types{\func{exchange}_i \; \gamma}{\func{exchange}_i \; \Gamma}{\allowbreak \func{exchange}_i \; P}{\func{exchange}_i \; \Xi}$.
\end{nitheorem}

\begin{proof}[Proof]
  All the above theorems are proved by induction on \textsc{Types} and \textsc{VarRef}.
  For details, refer to our mechanisation in Agda%
\begin{anonsuppress}
\cite{Zalakain2020Agda}
\end{anonsuppress}
.
\end{proof}

\subsection{Towards Subject Reduction}
\paragraph*{Subject Congruence}
This property states that applying structural congruence (\autoref{structural-congruence}) to a well-typed process preserves its well-typedness.
To prove this result, we must first introduce lemmas that establish that some syntactic manipulations can be inverted (\autoref{lm:lower-lift}, \autoref{lm:exchange-exchange}) and how unused variables relate to the preservation of leftovers (\autoref{lm:types-unused}).

@@

\begin{proof}[Proof]
  The proof is by induction on \textsc{Equals} $\eq{}$.
  Here we only consider those cases that are not purely inductive: the base cases for $\constr{struct}$ and their symmetric variants.
\begin{anonsuppress}
Full proof in \cite{Zalakain2020Agda}.
\end{anonsuppress}
  We proceed by induction on \textsc{StructCong} $\eqeq$:
  \begin{itemize}
    \item
      Case $\constr{comp-assoc}$: trivial, as leftover typing is naturally associative.
    \item
      Case $\constr{comp-sym}$ for $\comp{P}{Q}$: we use framing (\autoref{thm:framing}) to shift the output context of $P$ to the one of $Q$; and the input context of $Q$ to the one of $P$.
    \item
      Case $\constr{comp-end}$: trivial, as the typing rule for $\PO$ has the same input and output contexts.
    \item
      Case $\constr{scope-end}$: we show that the usage annotation of the newly created channel must be $\lz$, making the proof trivial.
      In the opposite direction, we instantiate the newly created channel to a type $\unit$ and a usage annotation $\lz$.
    \item
      Case $\constr{scope-ext}$ for $\new \comp{P}{Q}$: we to show that $P$ preserves the usage annotations of the unused variable (\autoref{lm:types-unused}) and then use strengthening (\autoref{thm:strengthening}).
      In the reverse direction, we use weakening (\autoref{thm:weakening}) on $P$ and show that lowering and then lifting $P$ results in $P$ (\autoref{lm:lower-lift}).
    \item
      Case $\constr{scope-comm}$:
      we use exchange (\autoref{thm:exchange}), and for the reverse direction exchange and \autoref{lm:exchange-exchange} to show that exchanging two elements in $P$ twice leaves $P$ unchanged. \qedhere
  \end{itemize}
\end{proof}

\paragraph*{Substitution}
This result is key to proving subject reduction.
In \autoref{thm:substitution-generalization} we prove a generalised version of substitution, where the substitition $\subst{P}{j}{i}$ is on any variable $i$.
Then, in \autoref{thm:substitution} we instantiate the generalised version to the concrete case where $i$ is the most recently introduced variable $\constr{0}$, as required by subject reduction.

@@

\begin{proof}[Proof]
  By induction on $\types{\gamma}{\Gamma_i}{P}{\Psi_i}$.
  Full proof can be found in \autoref{app:substitution-generalization}.
  \begin{itemize}
   \item
     Constructor $\PO$: observe that $\Gamma_i \equiv \Psi_i$ and thus $\Gamma_j \equiv \Psi_j$.
   \item
     Constructor $\new$: we proceed inductively.
   \item
     Constructors $\recv{}{}$, $\send{}{}$ and $\comp{}{}$: we must find $\Theta$ in \autoref{fig:subst} and split the diagram along its vertical axis to proceed by induction. \qedhere
 \end{itemize}
\end{proof}

@@

\begin{nitheorem}[Substitution]
  \label{thm:substitution}
  Let process $P$ be well typed in $\types{\gamma \comma t}{\Gamma \comma m}{P}{\Psi \comma \lz}$.
  Let $\contains{\gamma}{\Psi}{j}{t}{m}{\Xi}$.
  Then, we can substitute the variable references to $\constr{0}$ in $P$ with $\suc j$ so that the result is well typed in $\types{\gamma \comma t}{\Gamma \comma m}{\subst{P}{\suc j}{\constr{0}}}{\Xi \comma m}$.
\end{nitheorem}
\begin{proof}[Proof]
  For $\contains{\gamma}{\Gamma}{j}{t}{m}{\Theta}$ and $\types{\gamma \comma t}{\Theta \comma m}{P}{\Xi \comma \lz}$ for some $\Theta$, we use framing to derive them.
  Then, we use these to apply \autoref{thm:substitution-generalization}.
\end{proof}

@@

\paragraph*{Subject Reduction}
Finally we are ready to present our main result, stating  that if $P$ is well typed and it reduces to $Q$, then $Q$ is well typed.
The relation between the typing contexts used to type $P$ and $Q$ will be explained in \autoref{thm:subject-reduction}.
In the \picalc{} we distinguish between a reduction $P \reduce{\constr{internal}} Q$ on a channel internal to $P$, and a reduction $P \reduce{\constr{external} \; i} Q$ on a channel $i$ external to $P$ (refer to \autoref{operational-semantics}).
We first introduce an auxiliary lemma:

@@

\begin{proof}[Proof]
  By induction on $P \reduce{c} Q$. For the full details refer to our mechanisation in Agda.  
  \begin{itemize}
    \item
    Case $\constr{comm}$: we apply framing (\autoref{thm:framing}) (to rearrange the assumptions), substitution (\autoref{thm:substitution}) and strengthening (\autoref{thm:strengthening}).
